---
title: "lab4"
author: "Keven Lin"
date: 10/19/2021
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(readxl)
library(tidyverse)
library(ggplot2)
library(ggthemes)
defense <- read_excel('cyclonesFootball2020.xlsx', sheet='Defensive')
offensive <- read_excel('cyclonesFootball2020.xlsx', sheet='Offensive')
biography <- read_excel('cyclonesFootball2020.xlsx', sheet='Biography')
```

# Part one: Cleaning data


Although the three Excel sheets are loaded into R, they are not useful before some cleaning. Manipulate the three imported data frames:

1. The player names and opponent names should be factors.
```{r}
defense$Name = factor(defense$Name)
offensive$Name = factor(offensive$Name)
biography$Name = factor(biography$Name)

defense$Opponent_Opponent = factor(defense$Opponent_Opponent)
offensive$Opponent_Opponent = factor(offensive$Opponent_Opponent)
```
2. All offensive and defensive statistics should be numerical. The Weight column in biogrpahy should also be numeric. (Optional: Read ?across and learn to mutate multiple columns simultaneously using the across function in dplyr)
```{r}
defense = defense %>% mutate(across(Tackles_Solo:Pass_PB, ~as.numeric(.)))

offensive = offensive %>% mutate(across(c(Receiving_REC, Receiving_YDS, Receiving_TD, Rushing_ATT, Rushing_YDS, Rushing_TD, Passing_YDS, Passing_TD, Passing_INT), ~as.numeric(.)))

offensive$`Passing_CMP-ATT` = gsub("\r?\n|\r", " ", offensive$`Passing_CMP-ATT`)
offensive$`Passing_CMP-ATT` = gsub("\\s", "", offensive$`Passing_CMP-ATT`) 

offensive = offensive %>% mutate(Passing_CMP = as.numeric(substr(`Passing_CMP-ATT`, 1, 2)))
offensive = offensive %>% mutate(Passing_ATT = as.numeric(substr(`Passing_CMP-ATT`, 4, 5)))
offensive = within(offensive, rm(`Passing_CMP-ATT`))

biography$Weight = as.numeric(biography$Weight)
```

3. The Height column in biography should be numeric. (Hint: Use a unit that works well)
```{r}
biography = biography %>% mutate(Height_Feet = as.numeric(substr(Height, 1, 1)))
biography = biography %>% mutate(Height_Inch = as.numeric(substr(Height, 3, 4)))
biography = biography %>% mutate(Height = Height_Feet + (Height_Inch / 12))
biography = within(biography, rm(Height_Feet, Height_Inch))
# biography$Height
```

4. Add a column to both defense and offense indicating whether it is the first or the second game versus an opponent. We had a single game versus each opponent except for Oklahoma, which we played against twice. (Hint: You can use group_by(), mutate(), and row_number())

```{R}
offensive = offensive %>% group_by(Name) %>% mutate(GameNumber = ifelse( duplicated(Opponent_Opponent), 2, 1))
defense = defense %>% group_by(Name) %>% mutate(GameNumber = ifelse( duplicated(Opponent_Opponent), 2, 1))

str(offensive)


```
By cleaning the three data frames in R, at the end of this step you should have the following three clean data frames: defClean, offClean, and bioClean. Print the structures of these data frames using str().

```{R}
defClean = defense
offClean = offensive
bioClean = biography
str(defClean)
str(offClean)
str(bioClean)
```

# Part two: Tidying and exploring

For the following questions, work on the cleaned data frames defClean, offClean, and bioClean.

1. Reformat the offClean data frame into a tidy format using pivot_longer. The type of statistic (Receiving_REC for the number of receivings, Receiving_YDS for receiving yards, etc) is added as a new key column named stat.
```{r}
offClean = offClean %>%
  pivot_longer(!Name & !Opponent_Opponent, names_to = "stat", values_to = "values")
offClean
```
2. Obtain the player-level offensive statistics. Namely, for each player and each type of statistics, obtain the sum of the statistics in all games.
```{r}
offClean %>% group_by(Name, stat) %>% summarise(sum = sum(values))
```
3. Show the distributions of the player-level offensive statistics. The distribution of each offensive statistic should be shown in a separate facet. Explain the patterns you see.
```{r}

```
4. Did ISU have better offense against Oregon or Oklahoma? Answer this question by creating a scatterplot with the x- and y-axes being the receiving yards of the ISU players against each of these two opponents. A large number of receiving yards indicates good offense.

```{r}
oregon = offClean %>% filter(Opponent_Opponent %in% c('Oregon'), stat %in% c('Receiving_YDS'))
oklahoma = offClean %>% filter(Opponent_Opponent %in% c('Oklahoma'), stat %in% c('Receiving_YDS'))
oklahoma
ggplot(data = oregon, aes(x = values, y = values)) + geom_point()
ggplot(data = oklahoma, aes(x = values, y = values)) + geom_point()
```
### Yes, ISU scored more points against Oklahoma as the Median against Oklahoma is around 50 points compare to 18 points against Oregon. 

Now turn our attention to the player biography.

5. Separate the Hometown column into two columns, namely, the city and the state where the player comes from. Print the first few rows of these two columns. (Hint: look at the sep= argument of separate)

```{r}
bioClean = bioClean %>% separate(Hometown, c('City', 'State'))
bioClean %>% select(City, State)
```
6. How many players are there from each state?

```{R}
bioClean %>% group_by(State) %>% tally()
```
7. How does the performance of quarterback Brock Purdy relate to the performance of the ISU team? Comment on both the defense and offense.
```{r}
offClean %>% group_by(Name) %>% filter(Name %in% c("Purdy, Brock"), stat %in% c('Passing_CMP')) %>% summarise(Avg_Pass_Completed = mean(values, na.rm = TRUE))
offClean %>% group_by(Name) %>% filter(Name %in% c("Purdy, Brock"), stat %in% c('Passing_ATT')) %>% summarise(Avg_Pass_Attempted = mean(values, na.rm = TRUE))

offClean %>% group_by(stat) %>% filter(stat %in% c('Passing_CMP')) %>% summarise(Avg_Pass_Completed = mean(values, na.rm = TRUE))
offClean %>% group_by(stat) %>% filter(stat %in% c('Passing_ATT')) %>% summarise(Avg_Pass_Attempted = mean(values, na.rm = TRUE))
```
### when it comes to offensive, the action/performance of Brock Purdy will directly correspond to the team's performance. ###
### Purdy's average Pass completion is also the teams average of 20.25 throws. 
### Purdy's average pass attempted is also the teams average of 30.41667 throws. 

```{r}
defClean %>% group_by(Name) %>% filter(Name %in% c("Purdy, Brock")) %>% summarise(Avg_Tackles_Solo = mean(Tackles_Solo, na.rm = TRUE))
defClean %>% group_by(Name) %>% filter(Name %in% c("Purdy, Brock")) %>% summarise(Avg_Tackles_ASST = mean(Tackles_ASST, na.rm = TRUE))

mean(defClean$Tackles_Solo)
mean(defClean$Tackles_ASST)
```
### However, it is a different story for defense as Purdy's performance seems to have no effect on the performance of the team. 
### Purdy's average Tackles solo is 0.08333333 and average tackles assistant is 0. 
### This is wildly different from the teams where the average solo tackle is 0.9032258 and average tackle assistant is 0.7446237. 
### As we can see, Purdy hardly perform any tackles and never assisteted anyone. Compare to most invidual of the defensive team that performs almost 1 tackle per game and assist 0.75 tackles per game. 
